---
title: "aSDMs_Demo Application"
author: "Georgios Vagenas"
date: "2026-06-15"
output: html_document
---

### Start of Tutorial
aSDMs :: Freshwater SDMs – Complete Demo

This tutorial demonstrates the full **aSDMs workflow**:

- **Phase 1** – Global models for widespread species.
- **Phase 2** – Regional models for all species; for widespread ones the global prediction is used as an extra covariate.

The methodology is computational‑light (only 7 random species, few replicates) but exactly follows the production pipeline.  
All required files can be downloaded from: `https://saco.csic.es/s/Co8WNBa323ft3Qi`.

## Step 0 :: Required Packages
```{r}
# List of required packages

pkgs <- c("sf", "sdm", "terra", "dplyr", "tidyr", "mapview", "geodata", "raster",
          "duckdb", "arrow", "vandalico")
if (length(setdiff(pkgs, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(pkgs, rownames(installed.packages())))
}
# Load all packages silently
invisible(lapply(pkgs, library, character.only = TRUE))
```

## Step 1 :: Data Input & Species Filtering
```{r}


# set your working directory (adjust path)
setwd("path/to/your/project")

# 1.1 Occurrence data
iberia_occ <- vect("enriched_iberia_vect.shp")
global_occ <- vect("global_gbif_widespread.shp")

# 1.2 Read the species dataset, a set of points regional points is provided in the data folder (treat them as Endemics) but for full scale application global (GBIF) and regional (local databases) are needed.
species_list <- read.csv("enriched_species_grid_counts.csv")

# Keep species with ≥10 occurrences
adequate <- species_list %>% filter(Total_Grids >= 10)
robust_sp <- adequate$Sp

# subset vectors
iberia_occ <- iberia_occ[iberia_occ$Sp %in% robust_sp, ]
global_occ <- global_occ[global_occ$Sp %in% robust_sp, ]

# separate widespread vs. endemic
widespread_sp <- intersect(unique(iberia_occ$Sp), unique(global_occ$Sp))
endemic_sp    <- setdiff(unique(iberia_occ$Sp), widespread_sp)

cat("Widespread species:", length(widespread_sp), "\n")
cat("Endemic species:", length(endemic_sp), "\n")

# For demo, keep only a few species (optional)
set.seed(123)
demo_widespread <- sample(widespread_sp, 2)  # 2 widespread
demo_endemic    <- sample(endemic_sp, 5)      # 5 endemic
iberia_occ <- iberia_occ[iberia_occ$Sp %in% c(demo_widespread, demo_endemic), ]
global_occ <- global_occ[global_occ$Sp %in% demo_widespread, ]
widespread_sp <- demo_widespread
endemic_sp    <- demo_endemic
```

## Step 2 :: Predictors & Spatial Extents
```{r}

# 2.1 Predictor rasters
predictors_global   <- rast("predictors_finalized_global.tiff")
predictors_regional <- rast("predictors_finalized_regional.tiff")

# 2.2 Define predictor sets (must match raster layer names)
# Note: Regional Climate uses bio6_clima instead of bio17_clima (varies by region)
sets_global <- list(
  "Climate"       = c("bio5_clima", "bio16_clima", "bio17_clima", "bio15_clima", "bio4_clima"),
  "Hydroclimatic" = c("bio4_hydro", "bio1_hydro", "bio16_hydro", "bio17_hydro", "bio15_hydro")
)

sets_regional <- list(
  "Climate"            = c("bio5_clima", "bio16_clima", "bio6_clima", "bio15_clima", "bio4_clima"),
  "Hydroclimatic"      = c("bio4_hydro", "bio1_hydro", "bio16_hydro", "bio17_hydro", "bio15_hydro"),
  "Hydromorphological" = c("lka_pc_use", "dor_pc_pva", "sgr_dk_rav", "urb_pc_use", "for_pc_use")
)

# 2.3 Spatial extents, all the files are provided
ecoregions    <- vect("feow_hydrosheds.shp")
hydrosheds_H5 <- vect("HydroSHEDS_H5_merged.shp")  # all H5 basins
hydrosheds_H8 <- vect("HydroSHEDS_H8_merged.shp")
hydrosheds_H12<- vect("HydroSHEDS_H12_merged.shp")
study_area    <- vect("study_area_iberia.shp")

# 2.4 Training extent generation function
generate_extent <- function(occ, sp_name, eco, hydro, crop_area = NULL) {
  sp_occ <- occ[occ$Sp == sp_name, ]
  pts_eco <- terra::intersect(sp_occ, eco)
  eco_ids <- unique(pts_eco$FEOW_ID)
  sp_eco  <- eco[eco$FEOW_ID %in% eco_ids, ]
  pts_h   <- terra::intersect(sp_occ, hydro)
  h_ids   <- unique(pts_h$HYBAS_ID)
  sp_h    <- hydro[hydro$HYBAS_ID %in% h_ids, ]
  sp_poly <- crop(sp_h, sp_eco)
  sp_poly <- mask(sp_poly, sp_eco)
  sp_poly <- aggregate(sp_poly)
  if (!is.null(crop_area)) sp_poly <- crop(sp_poly, crop_area)
  return(sp_poly)
}


```

## Step 3 :: Build the Parquet Database (faster extraction, does not overloads memory)
```{r}
# 3.1 Global Parquet (for Phase 1)
global_parquet_file <- "aSDMs_Global_Parquet.parquet"

if (!file.exists(global_parquet_file)) {
  env_df <- as.data.frame(predictors_global, cells = TRUE, xy = TRUE, na.rm = TRUE)
  names(env_df) <- tolower(names(env_df))
  names(env_df) <- gsub("\\.", "_", names(env_df))
  write_parquet(env_df, global_parquet_file)
  cat("Global Parquet database created.\n")
} else {
  cat("Using existing Global Parquet file.\n")
}

# 3.2 Regional Parquet (for Phase 2) – contains hydromorphological variables
regional_parquet_file <- "aSDMs_Regional_Parquet.parquet"

if (!file.exists(regional_parquet_file)) {
  env_df_reg <- as.data.frame(predictors_regional, cells = TRUE, xy = TRUE, na.rm = TRUE)
  # ensure the column names match the raster layer names exactly
  layer_names <- names(predictors_regional)
  if (ncol(env_df_reg) == length(layer_names) + 2) {
    names(env_df_reg)[3:ncol(env_df_reg)] <- layer_names
  }
  names(env_df_reg) <- tolower(names(env_df_reg))
  names(env_df_reg) <- gsub("\\.", "_", names(env_df_reg))
  if (!"cell" %in% names(env_df_reg)) env_df_reg$cell <- 1:nrow(env_df_reg)
  write_parquet(env_df_reg, regional_parquet_file)
  cat("Regional Parquet database created.\n")
} else {
  cat("Using existing Regional Parquet file.\n")
}
```

## Step 4 :: Helper Function (Spatial Query + Background)
```{r}

# Generic function to fetch training data from a given Parquet file
get_train_data <- function(sp_name, sp_ext_poly, occ_df, t_vars, parquet_path, n_bg = NULL) {
  # 4.1 Get bounding box of the polygon
  sp_ext_coords <- as.vector(ext(sp_ext_poly))
  xmin <- sp_ext_coords['xmin']; xmax <- sp_ext_coords['xmax']
  ymin <- sp_ext_coords['ymin']; ymax <- sp_ext_coords['ymax']
  
  # 4.2 Query Parquet inside the bounding box
  con <- dbConnect(duckdb::duckdb())
  cols_sql <- paste(paste0("e.", t_vars), collapse = ", ")
  q <- sprintf("SELECT e.cell, e.x, e.y, %s FROM read_parquet('%s') e WHERE e.x BETWEEN %f AND %f AND e.y BETWEEN %f AND %f",
               cols_sql, parquet_path, xmin, xmax, ymin, ymax)
  box_data <- dbGetQuery(con, q)
  dbDisconnect(con)
  names(box_data) <- tolower(names(box_data))
  
  # 4.3 Mask to the exact polygon
  box_pts <- vect(box_data, geom = c("x", "y"), crs = crs(sp_ext_poly), keepgeom = TRUE)
  inside_pts <- box_pts[sp_ext_poly, ]
  poly_data <- as.data.frame(inside_pts)
  
  # 4.4 Presence points
  pres_ids <- occ_df[occ_df$Sp == sp_name, "cell_id"]
  valid_pres <- intersect(pres_ids, poly_data$cell)
  df_pres <- poly_data[poly_data$cell %in% valid_pres, ]
  df_pres$presence <- 1
  
  # 4.5 Background sampling (strictly from inside polygon)
  if (is.null(n_bg)) n_bg <- max(5, round(nrow(poly_data) * 0.05))
  set.seed(123)
  bg_cells <- sample(setdiff(poly_data$cell, valid_pres), n_bg, replace = FALSE)
  df_bg <- poly_data[poly_data$cell %in% bg_cells, ]
  df_bg$presence <- 0
  
  train_ready <- rbind(df_pres, df_bg)
  return(train_ready)
}


```
## Step 5 :: Prepare Output Structures
```{r}

global_output_dir <- "phase1_global"
regional_output_dir <- "phase2_regional"
dir.create(global_output_dir, recursive = TRUE)
dir.create(regional_output_dir, recursive = TRUE)

# Results table
results_all <- data.frame(
  species   = character(),
  phase     = character(),
  extent    = character(),
  set       = character(),
  AUC       = numeric(),
  COR       = numeric(),
  maxTSS    = numeric(),
  maxKappa  = numeric(),
  CBI       = numeric(),
  stringsAsFactors = FALSE
)

algo_list <- c('glm', 'brt', 'rf')
n_reps_demo <- 2   # low for speed (production uses 5)
ref_raster <- predictors_regional[[1]]

```

## Step 6 :: Phase 1 – Global Models (widespread species)
```{r}

extents_global <- list(eco = ecoregions, H5 = hydrosheds_H5, H8 = hydrosheds_H8, H12 = hydrosheds_H12)

for (sp in widespread_sp) {
  clean_sp <- gsub(" ", "_", sp)
  sp_global_occ <- as.data.frame(global_occ[global_occ$Sp == sp, ])
  # compute cell IDs using the global reference raster
  sp_global_occ$cell_id <- cellFromXY(predictors_global[[1]], crds(global_occ[global_occ$Sp == sp, ]))
  
  for (ext_name in names(extents_global)) {
    ext_poly <- generate_extent(global_occ, sp, ecoregions, extents_global[[ext_name]])
    if (is.null(ext_poly) || nrow(ext_poly) == 0) next
    
    for (set_name in names(sets_global)) {
      cat(sprintf("GLOBAL | %s | %s | %s\n", sp, ext_name, set_name))
      out_path <- file.path(global_output_dir, clean_sp, ext_name, set_name)
      dir.create(out_path, recursive = TRUE, showWarnings = FALSE)
      
      t_vars <- sets_global[[set_name]]
      # Use global Parquet for Phase 1
      train_ready <- get_train_data(sp, ext_poly, sp_global_occ, t_vars, 
                                    parquet_path = global_parquet_file)
      
      d <- sdmData(as.formula(paste0("presence ~ ", paste(t_vars, collapse = "+"), " + coords(x+y)")),
                   train = train_ready[train_ready$presence == 1, ],
                   bg    = train_ready[train_ready$presence == 0, ])
      
      m <- tryCatch({
        sdm(presence ~ ., d, methods = algo_list, replication = 'boot', n = n_reps_demo)
      }, error = function(e) { NULL })
      if (is.null(m)) next
      
      p_ens <- ensemble(m, train_ready[, c("x", "y", t_vars)],
                        setting = list(method = 'weighted', stat = 'AUC'))
      res_df <- data.frame(x = train_ready$x, y = train_ready$y, val = as.numeric(p_ens[[1]]))
      global_raster <- rast(res_df, type = "xyz", crs = crs(ext_poly))
      
      # crop to Iberia and save
      iberia_masked <- mask(crop(global_raster, study_area), study_area)
      writeRaster(iberia_masked, file.path(out_path, "ensemble_global_iberia.tif"), overwrite = TRUE)
      
      # evaluation
      e <- evaluates(d, p_ens)
      b <- sdm:::.boyce(e@observed, e@predicted)
      metrics <- data.frame(
        species = sp, phase = "global", extent = ext_name, set = set_name,
        AUC = as.numeric(e@statistics$AUC[1]),
        COR = as.numeric(e@statistics$COR[1]),
        maxTSS = max(e@threshold_based$TSS),
        maxKappa = max(e@threshold_based$Kappa),
        CBI = ifelse(!is.null(b$CBI), as.numeric(b$CBI[1]), NA)
      )
      results_all <- rbind(results_all, metrics)
      write.csv(metrics, file.path(out_path, "eval.csv"), row.names = FALSE)
      
      # variable importance
      vi <- getVarImp(m, id = "ensemble")
      write.csv(vi, file.path(out_path, "varImp.csv"))
      
      rm(d, m, global_raster, iberia_masked, p_ens); gc()
    }
  }
}


```

## Step 7 :: Phase 2 – Regional Models (all regional species)
```{r}
extents_regional <- list(eco = ecoregions, H5 = hydrosheds_H5, H8 = hydrosheds_H8, H12 = hydrosheds_H12)

# Pre‑compute cell IDs for Iberian occurrences (using regional reference raster)
iberia_occ_df <- as.data.frame(iberia_occ)
iberia_occ_df$cell_id <- cellFromXY(ref_raster, crds(iberia_occ))

for (sp in c(widespread_sp, endemic_sp)) {
  clean_sp <- gsub(" ", "_", sp)
  sp_occ <- iberia_occ_df[iberia_occ_df$Sp == sp, ]
  
  for (ext_name in names(extents_regional)) {
    ext_poly <- generate_extent(iberia_occ, sp, ecoregions, extents_regional[[ext_name]], study_area)
    if (is.null(ext_poly) || nrow(ext_poly) == 0) next
    
    for (set_name in names(sets_regional)) {
      cat(sprintf("REGIONAL | %s | %s | %s\n", sp, ext_name, set_name))
      out_path <- file.path(regional_output_dir, clean_sp, ext_name, set_name)
      dir.create(out_path, recursive = TRUE, showWarnings = FALSE)
      
      t_vars <- sets_regional[[set_name]]
      
      # Use regional Parquet for Phase 2
      train_ready <- get_train_data(sp, ext_poly, sp_occ, t_vars, 
                                    parquet_path = regional_parquet_file)
      
      # inject global suitability if species is widespread
      if (sp %in% widespread_sp) {
        global_ras <- rast(file.path(global_output_dir, clean_sp, ext_name,
                                     ifelse(set_name %in% c("Climate","Hydroclimatic"),
                                            set_name, "Hydroclimatic"),
                                     "ensemble_global_iberia.tif"))
        if (file.exists(global_ras)) {
          global_vals <- terra::extract(global_ras, cbind(train_ready$x, train_ready$y))[, 2]
          train_ready$global_suit <- ifelse(is.na(global_vals), 0, global_vals)
          t_vars <- c(t_vars, "global_suit")
        }
      }
      
      d <- sdmData(as.formula(paste0("presence ~ ", paste(t_vars, collapse = "+"), " + coords(x+y)")),
                   train = train_ready[train_ready$presence == 1, ],
                   bg    = train_ready[train_ready$presence == 0, ])
      
      m <- tryCatch({
        sdm(presence ~ ., d, methods = algo_list, replication = 'boot', n = n_reps_demo)
      }, error = function(e) { NULL })
      if (is.null(m)) next
      
      p_ens <- ensemble(m, train_ready[, c("x", "y", t_vars)],
                        setting = list(method = 'weighted', stat = 'AUC'))
      reg_raster <- rast(data.frame(x = train_ready$x, y = train_ready$y, val = as.numeric(p_ens[[1]])),
                         type = "xyz", crs = crs(ext_poly))
      writeRaster(reg_raster, file.path(out_path, "ensemble_regional.tif"), overwrite = TRUE)
      
      e <- evaluates(d, p_ens)
      b <- sdm:::.boyce(e@observed, e@predicted)
      metrics <- data.frame(
        species = sp, phase = "regional", extent = ext_name, set = set_name,
        AUC = as.numeric(e@statistics$AUC[1]),
        COR = as.numeric(e@statistics$COR[1]),
        maxTSS = max(e@threshold_based$TSS),
        maxKappa = max(e@threshold_based$Kappa),
        CBI = ifelse(!is.null(b$CBI), as.numeric(b$CBI[1]), NA)
      )
      results_all <- rbind(results_all, metrics)
      write.csv(metrics, file.path(out_path, "eval.csv"), row.names = FALSE)
      
      vi <- getVarImp(m, id = "ensemble")
      write.csv(vi, file.path(out_path, "varImp.csv"))
      
      rm(d, m, reg_raster, p_ens); gc()
    }
  }
}


```

## Step 8 :: Final Summary
```{r}
write.csv(results_all, "full_evaluation_summary.csv", row.names = FALSE)
cat("Pipeline completed. Results saved to:", global_output_dir, "and", regional_output_dir, "\n")
cat("Evaluation summary written to full_evaluation_summary.csv\n")

```
### End of Tutorial
