---
title: "aSDM_Toy Problem Application"
author: "Georgios Vagenas"
date: "2025-10-26"
output: html_document
---

### Start of Tutorial
aSDMs :: Aquatic Species Distribution Models - Tutorial

This tutorial demonstrates a workflow for building aquatic Species Distribution Models (aSDMs) using occurrence data and environmental predictors. In this exercise, we provide a computational-light exercise with seven random species that demonstrates our methodology step-by-step, by implementing the pre-constrained h5 climate aSDMs, thus making it accessible for readers to replicate and explore. To download the repository which includes all the required files for the toy problem execution download from: https://saco.csic.es/s/Co8WNBa323ft3Qi.

## Step 0 :: Required Packages
```{r}
# List of required packages
pkgs <- c("sf", "sdm", "dismo", "dplyr", "tidyr", 
          "mapview", "geodata", "raster", "RColorBrewer",
          "terra", "usdm", "randomForest", "parallel")

# Install missing packages in one command
if (length(setdiff(pkgs, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(pkgs, rownames(installed.packages())))
}

# Load all packages silently
invisible(lapply(pkgs, library, character.only = TRUE))
```

## Step 1 :: Data Input
```{r}
setwd("XXX/XXX/XXX")
data_sp<-vect("XXX/random_species_subset_sp10.shp")
terra_df <- data_sp  # Full occurrence dataset
unique_species <- unique(terra_df$species_id)
numbers_sp <- length(unique_species)
numbers_sp
```

## Step 2 :: Hydroclimatic Predictors
```{r}
riveratl_global <- rast("XXX/top5_global_rivers_atlas_raster_list_30sec.tif")
bioclim_global_rn <- rast("XXX//bioclim_global_rn_5vars_30sec.tif")

```

## Step 3 :: Initialize SDM Structures
```{r}
results_df <- data.frame(
  id = integer(),
  species_name = integer(),
  e_AUC = numeric(),
  e_COR = numeric(),
  t_maxSSS = numeric(),
  t_maxkappa = numeric(),
  t_prevalence = numeric(),
  CBI = numeric(),
  maxKappa = numeric(),
  maxTSS = numeric(),
  obs_prevalence = numeric(),
  stringsAsFactors = FALSE
)

raster_list <- list()
combined_rasters <- rast()
sdm_d <- list()
```

## Step 4 :: Spatial Processing - Allocation to watersheds
```{r}
watershed <- vect("XXX/H5_Iberian/H5_Iberian.shp") #the user should define the spatial layer, in our files the H8 and H12 alternatives are available
 #the user should define the spatial layer, in our files the H8 and H12 alternatives are available

# CRS Handling and Conversion
if (is.factor(terra_df$species_id)) {
  terra_df$species_id <- as.numeric(levels(terra_df$species_id))[terra_df$species_id]
} else {
  terra_df$species_id <- as.numeric(terra_df$species_id)
}
crs(terra_df) <- "EPSG:4326"
miteco_sp_vect <- terra_df

# Intersect occurrences with watersheds
miteco_sp_in <- terra::intersect(miteco_sp_vect, watershed)

# Data Conversion Pipeline
data <- as.data.frame(miteco_sp_in)
data$geometry <- NULL
coords <- geom(miteco_sp_in)
terra_df_o <- cbind(coords, data)

# Species Filtering (≥10 obs)
terra_df_or <- terra_df_o %>%
  mutate(species_id = as.numeric(factor(terra_df_o$species_id)))

species_name_id <- dplyr::select(terra_df_or, "species_id", "HYBAS_ID")

terra_df_filtered <- terra_df_or %>%
  group_by(species_id) %>%
  filter(n() >= 10) %>%
  ungroup() %>%
  droplevels()


# First, convert your spatial vector data to a dataframe
terra_df_conv <- as.data.frame(terra_df_filtered)

# Now create the presence-absence matrix
presence_absence_matrix <- terra_df_conv %>%
  mutate(presence = 1) %>%
  dplyr::select(species_id, HYBAS_ID, presence) %>%
  distinct() %>%
  pivot_wider(names_from = HYBAS_ID, values_from = presence, values_fill = list(presence = 0)) %>%
  as.data.frame()

# Order ascendingly by species_id
presence_absence_matrix <- presence_absence_matrix[order(presence_absence_matrix$species_id),]

# Convert presence-absence table back to long format
presence_long <- presence_absence_matrix %>%
  pivot_longer(-species_id, names_to = "HYBAS_ID", values_to = "presence")


# Spatial Template Preparation
iberian_watersheds <- watershed
aggregated_vec <- aggregate(iberian_watersheds)
resolution <- res(bioclim_global_rn)
raster_template <- rast(ext(aggregated_vec), resolution = resolution)
aggregated_raster <- rasterize(aggregated_vec, raster_template, field = 1, fun = "count")
aggregated_raster[!is.na(aggregated_raster)] <- NA
crs(aggregated_raster) <- "EPSG:4326"


combined_rasters<-resample(combined_rasters,aggregated_raster)
crs(combined_rasters)<- "EPSG:4326"

```
## Step 5 :: aSDM Models Prediction - Fitting
```{r}
input_cov <- bioclim_global_rn  # Primary covariates climate aSDM - in case the user wants to apply hydrological or hydroclimatic or hierarchical approaches the fundamental documentation should be followed

# Initialize lists
polygon_list <- list()
raster_list <- list()
sdm_d <- list()
combined_rasters_stack <- rast()
results_df <- data.frame()

# Define the vector of numbers
numbers_sp <- levels(as.factor(terra_df$species_id))


for (i in 1:length(numbers_sp)) {
  
  #set covariates
  input_cov <- bioclim_global_rn  # Use your primary covariates
  
  # Get presence data for the current species
  species_data <- presence_long %>%
    filter(species_id == unique(presence_long$species_id)[[i]] & presence == 1)
  
  # Get the watershed IDs
  watershed_id <- species_data$HYBAS_ID
  
  # Get the corresponding freshwater polygon for this HYBAS_ID
  masked_watershed <- watershed[watershed$HYBAS_ID %in% watershed_id, ]
  
  # Connect the watersheds produced in a single polygon
  polygon_list[[i]] <- masked_watershed
  
  # Dissolve the different watersheds to function as the training region
  dissolve <- aggregate(polygon_list[[i]], dissolve = TRUE)
  
  # Combine the raster objects
  combined_raster <- c(input_cov)
  
  ############ RUN THE SDMs #############
  
  bioc_gal_in <- combined_raster
  
  crs(dissolve) <- "EPSG:4326"
  
  # Crop the bioclim to the extent of the watersheds where the species belongs to
  bioc_gal_in_crop <- crop(bioc_gal_in, dissolve, mask = TRUE)
  
  # Pre-set the dataset - CORRECTED: use single brackets [i]
  terra_df_demo <- terra_df[terra_df$species_id == numbers_sp[i], ]
  terra_df_demo$species_id[terra_df_demo$species_id == numbers_sp[i]] <- 1
  
  # Convert SpatVector to data frame with coordinates
  terra_df_demo_df <- as.data.frame(terra_df_demo, geom = "XY")
  
  # Create SpatialPointsDataFrame from coordinates
  terra_df_demo_f <- SpatialPointsDataFrame(
    coords = terra_df_demo_df[, c("x", "y")],
    data = data.frame(species_id = rep(1, nrow(terra_df_demo_df))),
    proj4string = CRS("+init=EPSG:4326")
  )
  
  # Generate background points equal to 5% of the training area
  background_points <- sum(freq(bioc_gal_in_crop[[1]]))* 0.05
  
  # Convert SpatRaster to RasterBrick
  r_bioc_gal_in_crop <- as(bioc_gal_in_crop, "Raster")
  
  d <- sdmData(species_id ~ ., terra_df_demo_f, predictors = r_bioc_gal_in_crop, 
               bg = list(method = 'gRandom', n = round(background_points), exclude = TRUE))
  
  # Store the background points in a vector to be evaluated at a later stage
  sdm_d[[i]] <- d
  
  # SDM function to fit the models
  m <- sdm(species_id ~ ., d, methods = c('glm', 'brt', 'rf'), replication = c('boot'),
           test.p = 30, n = 2, parallelSetting = list(ncore = 4, method = 'parallel'))
  
  # Current prediction to the rest of the 30%
  p2 <- predict(m, r_bioc_gal_in_crop)
  
  en1 <- ensemble(m, p2, setting = list(method = 'weighted', stat = 'auc'))
  e <- evaluates(d, en1)
  bc <- sdm:::.boyce(e@observed, e@predicted)
  
  results_df <- rbind(results_df, data.frame(
    id = i,
    species_id = paste("Species", i),
    e_AUC = e@statistics$AUC,
    e_COR = e@statistics$COR[1],
    CBI = bc$CBI,
    maxTSS = e@threshold_based$TSS[2],
    maxKappa = e@threshold_based$Kappa[5],
    t_maxSSS = e@threshold_based$threshold[2],
    t_maxkappa = e@threshold_based$threshold[5],
    t_prevalence = e@threshold_based$threshold[10],
    obs_prevalence = length(d@species$species_id@presence) / 
      (length(d@species$species_id@background) + length(d@species$species_id@presence))
  ))
  
  
  #name the raster file based on the species name
  species_name <- terra_df$Species[terra_df$species_id == i][1]
  names(en1) <- species_name
  
  #Combine the aSDM into the Iberian extent
  crop_en1<-mask(en1,aggregated_vec)
 
  mask_en1<- resample(crop_en1, combined_rasters, method = "near")
  #plot(mask_en1)
  
  # Reproject raster1 to the CRS of raster2
  crs(mask_en1) = "EPSG:4326"
  
  #store all the species in a combon raster file
  combined_rasters_stack <- c(combined_rasters_stack,mask_en1)
  
  # Define the file name for saving and NAME IT BASE ON THE SPECIES ACCORDING TO THE INTIAL LEDGER
  file_name <- paste0("deleteplease.tif")
  
  # Save the raster file
  writeRaster(en1, filename = file_name, overwrite = TRUE)
  
  # Print a message indicating the file has been saved
  cat("Saved:", file_name, "for iteration:", i, "\n")
}

# Assess the number of layers and the generated habitat suitability raster layer
nlyr(combined_rasters_stack)

plot(combined_rasters_stack)

```

## Step 6 :: Post-Processing - Thresholding based on prevalence
```{r}
# Threshold Application
thresholded_rasters_list <- list()

for (i in 1:max(results_df$id)) {
  threshold_value <- results_df$obs_prevalence[i]
  thresholded_raster <- app(
    combined_rasters_stack[[i]], 
    fun = function(x) ifelse(x > threshold_value, 1, 0)
  )
  thresholded_rasters_list[[i]] <- thresholded_raster
}

# Combine all at once
thresholded_rasters_comb <- rast(thresholded_rasters_list)
names(thresholded_rasters_comb) <- results_df$species_id[1:length(thresholded_rasters_list)]

# Species Richness Calculation
masked_rasters <- lapply(thresholded_rasters_comb, function(r) {
  r[is.na(r)] <- 0
  return(r)
})
stacked_raster <- rast(masked_rasters)
final_stack <- sum(stacked_raster)

#Plot species richness based on thresholded prevalence per species
plot(stacked_raster)

#Plot species richness based on thresholded prevalence for all species summed
plot(final_stack)

# Final Outputs
writeRaster(combined_rasters_stack, "preh5_clima_endemics_1km.tif", overwrite = TRUE)
writeRaster(final_stack,"species_richness_thresholded_1km.tif",overwrite=TRUE)
write.csv(results_df, "preh5_clima_endemics_metrics.csv", row.names = FALSE)
```
### End of Tutorial